---
title: "Examples"
author: "Sebastian Kopf"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Load **isocyclr** and other packages used in these examples. Help on all functions is available within R (e.g. via `?add_standard_reaction`).

```{r, message=FALSE, warning=FALSE}
library(isocyclr)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
```

## Set up reaction path

Set up a reaction network akin to Figure 2 in John Hayes' classic *Fractionation of carbon and hydrogen isotopes in biosynthetic processes. Rev Mineral Geochem 43, 225–277. (2001)*.

```{r}
path <- isopath() %>% 
  add_isotope("carbon") %>% 
  add_component("AA", carbon, variable = FALSE) %>% 
  add_component(LETTERS[2:9], carbon) %>% # add B through I
  add_component(c("DD", "GG", "II"), carbon, variable = FALSE) %>% # out flux
  add_standard_reaction(AA == B, eps.carbon = e1, flux = dm) %>% 
  add_standard_reaction(B == C, eps.carbon = e2, flux = dm) %>% 
  add_standard_reaction(C == D, eps.carbon = e3, flux = f3 * dm) %>% 
  add_standard_reaction(D == DD, eps.carbon = 0, flux = f3 * dm, name = "D out") %>% 
  add_standard_reaction(C == E, eps.carbon = e4, flux = (1-f3) * dm) %>% 
  add_standard_reaction(E == F, eps.carbon.eq = e5, flux = (1-f3) *dm, eq_ratio = "P/S") %>%
  add_standard_reaction(F == G, eps.carbon = e6, flux = (1-f3) *f6 * dm) %>% 
  add_standard_reaction(G == GG, eps.carbon = 0, flux = (1-f3) *f6 * dm, name = "G out") %>% 
  add_standard_reaction(F == H, eps.carbon = e7, flux = (1-f3) * (1-f6) * dm) %>% 
  add_standard_reaction(H == I, eps.carbon = e8, flux = (1-f3) * (1-f6) * dm) %>% 
  add_standard_reaction(I == II, eps.carbon = 0, flux = (1-f3) * (1-f6) * dm, name = "I out") 
```

#### Schematic

```{r rxn_diagram, fig.width = 8, fig.height = 5}
path %>% generate_reaction_diagram() + coord_equal()
```

#### System of differential equations

The `get_ode_matrix()` makes it easy to look at the system of differential equations generated for the reaction network. To test what it evaluates  to for the first step with the given parameters, take a look at the `evaluate=TRUE` parameter in `?get_ode_matrix` (requires parameters to be set so the variables can all be evaluated).

```{r}
path %>% get_ode_matrix() %>% knitr::kable()
```


## Assign parameters

The symbols used in reaction network setup can be assigned numeric values (multiple scenarios if needed) that get evaluated upon running the model.

```{r}
path <- path %>% 
  set_parameters(
    data_frame(
      scenario = c("low flux", "high flux"),
      # fluxes and flux fractions
      dm = c(0.1, 1), f3 = 0.2, f6 = 0.2,
      # isotopic effects
      e1 = 0, e2 = 25, e3 = 15, e4 = 35, e5 = 8, e6 = 20, e7 = 0, e8 = 14,
      # starting isotopic composition
      AA.carbon = 0, B.carbon = 0, C.carbon = 0, D.carbon = 0, E.carbon = 0, 
      F.carbon = 0, G.carbon = 0, H.carbon = 0, I.carbon = 0,
      # pool sizes for variable components
      B = 10, C = 1, D = 20, E = 10, F = 50, G = 20, H = 10, I = 1
      )
  ) 
```

## Run model

Running the system of differential equations is easily done and uses the ode solvers of the **deSolve** package.

```{r}
model <- path %>% run_model(time_steps = 500)
```

## Plot time course

```{r time_course, fig.width = 8, fig.height = 6}
model %>% 
  gather(reservoir, delta, ends_with("carbon")) %>%
  ggplot() + aes(time, delta, color = reservoir) +
  geom_line() + theme_bw() +
  labs(y = expression(delta*13*'C')) +
  facet_grid(~scenario)
```

## Run model with event

Additional parameters can be passed and will be forwarded to the ode solver, e.g. introducing different starting isotopic composition part way through via a timed event.

```{r time_course_event, fig.width = 8, fig.height = 6}
model2 <- path %>% run_model(
  time_steps = 500, make_state_var = c("AA.carbon"),
  events = list(data = data.frame(
    var = "AA.carbon", time = 250, value = 10, method = "rep")))
last_plot() %+% 
  gather(model2, reservoir, delta, ends_with("carbon"))
```

## Run to steady state

The model can also be run to steady-state using runsteady functionality from the *rootSolve* package. All parameters are forwarded to the ode/root solver.

```{r}
steady <- path %>% run_steady_state(stol = 1e-5, rtol = 1e-3)
```

#### Steady state isotopic composition

Looking at the steady state results, there is no difference between low and high flux once steady-state is reached (the correct behavior). Since fractionation is calculated precisely when using `add_standard_reaction()` (without any $\alpha \approx 1$ approximations), the isotopic offsets are not exactly the $\epsilon$ values provided.

```{r steady_state, fig.width = 8, fig.height = 6}
path %>% generate_reaction_diagram(
  steady %>% gather(component, y, ends_with("carbon")) %>% 
    mutate(component = sub(".carbon", "", component, fixed = T)) %>% 
    select(scenario, component, y)
) + facet_grid(~scenario)
```

Changing the the branching flux at nodes however, does change the outcome as expected. As an example:

```{r}
steady2 <- path %>% 
  set_parameters(scenario = c("low D/E branching ratio", "high D/E branching ratio"),
                 dm = c(1, 1), f3 = c(0.2, 0.9)) %>% 
  run_steady_state(stol = 1e-5, rtol = 1e-3)
```

```{r steady_state_branching, fig.width = 8, fig.height = 6}
path %>% generate_reaction_diagram(
  steady2 %>% gather(component, y, ends_with("carbon")) %>% 
    mutate(component = sub(".carbon", "", component, fixed = T)) %>% 
    select(scenario, component, y)
) + facet_grid(~scenario)
```

